---
title: "Quantitative Methods and Statistics"
subtitle: "Exam - Winter Term 2025/26"
author:
  - name: "Jakob Hoffmann"
    affiliation: "Economic Geography Group, LMU Munich"
format:
  html:
    toc: true
editor: 
  markdown: 
    wrap: sentence
---

# Introduction

A key measure of a country's integration into the global economy is **trade openness**, calculated as the sum of exports and imports divided by GDP (expressed as a percentage).
A central question in understanding globalization is whether more open economies tend to be richer.
The examination of this question is the primary subject of this exam.

The exam has three parts with a sum of **30 points**:

1.  Loading and wrangling data (10 points)

2.  Describing and visualizing data (10 points)

3.  Fitting a model to data (10 points)

# Problem 1: Load and wrangle data (10 points)

*HINT:* Don't forget to load the packages you need before starting with the analysis.

The file `trade_raw.csv` in the folder `data/exam` contains data about trade openness (trade as % of GDP) obtained from Our World in Data.

1)  Load the file into a data frame and save it to a variable called `trade_raw`. (*1 points*)

```{r}
library(tidyverse)
trade_raw <- read_csv("data/exam/trade_raw.csv")
```

2)  Inspect the data frame. How many rows and columns does it have? What are the names of its variables? (*1 points*)

```{r}

```

3)  Write a piece of code that keeps only the columns for country name, country code, year, and trade openness and renames them to `country`, `code`, `year`, and `trade_openness`.
    (*1 points*)

    HINT: Remember to put variable names containing spaces into backticks (\`variable name\`).

```{r}

select(
  trade_raw,
  country = Entity,
  code = Code,
  year = Year,
  trade = `Trade (% of GDP)`
)
```

4)  The data set contains non-country entities (such as world aggregates and regions). Write a piece of code that keeps only actual countries by using the fact that ISO country codes have exactly three characters. (*1 points*)

```{r}
filter(
  trade_raw,
  nchar(Code) == 3
)
```

5)  Write a piece of code that keeps only observations for the year 2019. (*1 points*)

```{r}
filter(
  trade_raw,
  Year == 2019
)
```

6)  Write a pipeline that performs the three previous steps in sequence and store the result in a variable called `trade`. (*2 points*)

```{r}
trade <- trade_raw |>
  select(
    country = Entity,
    code = Code,
    year = Year,
    trade = `Trade (% of GDP)`
  ) |>
  filter(nchar(code) == 3, year == 2019)
```

7)  There is a second data set called `gdp_raw.csv` containing GDP per capita data. Load it, inspect it, and apply the same steps from 3 to 5 in one pipeline (make sure to name variables appropriately). Save the result as a variable called `gdp`. (*2 points*)

```{r}
gdp <- read_csv("data/exam/gdp_raw.csv") |>
  select(
    country = Entity,
    code = Code,
    year = Year,
    gdp = `GDP per capita`
  ) |>
  filter(nchar(code) == 3, year == 2019)
```

8)  Combine the two data frames using an inner join. (*1 points*)

```{r}
trade_gdp <- inner_join(gdp, trade)
```

# Problem 2: Describe and visualize the data (10 points)

The file `trade_gdp.csv` contains a pre-cleaned data set with trade openness, GDP per capita, a variable with a democracy index, and world region for 153 countries in 2019.
Load this file into a variable called `df` and use it for the remaining problems.

```{r}
df <- read_csv("data/exam/trade_gdp.csv")
```

1)  GDP per capita is highly skewed. Create a new variable `log_gdp_pc` containing the natural logarithm of GDP per capita. Show the distribution of the resulting variable as a histogram. (*2 points*)

    *HINT*: Remember to overwrite your data frame if you want changes to persist.

```{r}
df <- mutate(df, log_gdp_pc = log(gdp_per_capita))
ggplot(df, aes(log_gdp_pc)) + geom_histogram()
```

2)  Create a scatter plot showing the relationship between trade openness (x-axis) and log GDP per capita (y-axis). Color the points by world region and describe the result in one sentence. (*2 points*)

```{r}
ggplot(df, aes(trade_openness, log_gdp_pc, color = region)) + geom_point()
```

3)  Compute the Pearson correlation coefficient between trade openness and log GDP per capita. Interpret the result in one sentence. (*2 points*)

```{r}
cor(df$trade_openness, df$log_gdp_pc)
```

4)  Compute the mean, median, and standard deviation for trade openness by world region and sort the resulting data frame by the mean (descending). Give a brief summary of the result. (*2 points*)

```{r}
group_by(df, region) |>
  summarize(mean = mean(trade_openness)) |>
  arrange(desc(mean))
```

5)  Visualize the distribution of trade openness by world region (pick an adequate form of visualization). (*2 points*)

```{r}
ggplot(df, aes(region, trade_openness)) + geom_boxplot()
```

# Problem 3: Fit a model to the data (10 points)

1)  Fit a simple linear regression model predicting log GDP per capita from trade openness. Save the model as `m1`. (*1 points*)

```{r}
m1 <- lm(log_gdp_pc ~ trade_openness, data = df)
m1

library(rstanarm)
m1 <- stan_glm(log_gdp_pc ~ trade_openness, data = df)
```

2)  Visualize the linear model estimated above against the scatterplot of the data using `ggplot`. Use the visualization to get a (rough) prediction of log GDP per capita for a country with a trade openness of 200%. (*1 points*)

```{r}
ggplot(df, aes(trade_openness, log_gdp_pc)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

3)  Give an interpretation of the coefficient for trade openness.
    (*1 points*)

    *HINT*: If you're using `stan_glm()` and want to see more digits in the printed model result, you can call `print(model, digits = 5)`.

```{r}

print(m1, digits = 5)
```

4)  Summarize the estimation uncertainty. Is the coefficient for trade openness greater than zero with high certainty? (*1 points*)

```{r}

```

5)  The data set also contains a **democracy index** (ranging from 0 to 1, where higher values indicate more democratic institutions). Give a short discussion about why this might (not) be a confounder for the causal effect of trade openness on GDP per capita. (*2 points*)

```{r}

```

6)  Estimate a multiple regression model controlling for the level of democratic institutions. Give a brief interpretation of the coefficient of trade openness in this model. (*2 points*)

```{r}

m2 <- stan_glm(log_gdp_pc ~ trade_openness + democracy, data = df)
m2
```

7)  Compare this model to the first one and summarize the differences. How did adjusting for the democracy index change the estimated effect of trade openness? (*2 points*)

```{r}
print(m2, digits = 5)
```
