---
title: "Quantitative Methods and Statistics"
subtitle: "An applied course using the `R` programming language"
author: 
  - name: "Jakob Hoffmann"
    affiliation: "Economic Geography Group, LMU Munich"
format:
  html:
    toc: true
---

# Modeling statistical relationships

In the last sessions, we have seen how to use probability to both describe randomness in the data generating process we ultimately care about _and_ for statistical inference about unknown quantities (i.e., parameters) in the model.
These are the most basic bulding blocks of statistical modeling. However, so far the models we have looked at were not particularly interesting.

Just as with descriptive statistics, statistical modeling starts to shine once we look at multiple variables and their relationships with each other.
Typically, variables play different roles in a statistical model: Often there is a primary variable of interest, the variation of which we want to understand.
This is called the **outcome** or dependent variable.
The purpose of the models we build is then to describe how the outcome varies with respect to one or more **predictors** or independent variables.
With this setup, we focus our attention on modelling the DGP of the outcome while the predictor variables are typically taken for granted and not modeled explicitly.


# Linear regression

The probably most widely used model of a statistical relationship is _linear regression_, which models an outcome $y$ as a function of a predictor $x$.
As before, we don't presume to fully understand the variation in the outcome $y$ and thus model it as random. 
The most common choice is the normal distribution (which we have seen before):

$$
y \sim \textrm{Normal}(\mu, \sigma)
$$

This time, however, we treat $\mu$ (the mean of the distribution) not just as a parameter to be set or estimated, but as a function of a predictor $x$.
The simplest functional form of the relationship is just a line:

$$
\mu = \alpha + x \beta
$$

Instead of just two parameters, this model now has three:
The line parameters $\alpha$ (the _intercept_) and $\beta$ (the _slope_), which together determine the mean of the outcome $y$, conditional on $x$x), and the standard deviation $\sigma$, which determines the amount of random variation around the mean.

## Simulating linear regression

As before, this is a _generative_ model of a _data generating process_ and so the best way to see it in action is to simulate data from it.
To do so, we need to set values for the thee parameters and for the predictor $x$, which, remember, we assume to be given:

```{r}
# The model parameters
alpha <- 2
beta <- 1
sigma <- 0.5

# how many data points to simulate
n <- 100

# The predictors values for each of the n data points
x <- seq(0, 4, length.out = n)
```

With these values, we can simulate a $y_i$ for each of the $n$ predictor values $x_i$:


```{r}
mu <- alpha + x * beta
y = rnorm(n, mean = mu, sd = sigma)
```

Let's plot these simulated data alongside the line we just specified to get a better sense of what just happened:


```{r, warning = FALSE, message = FALSE}
library(tidyverse)

ggplot(tibble(x, y), aes(x, y)) +
  geom_point() +
  geom_abline(
    intercept = alpha,
    slope = beta,
    color = "tomato"
  )
```

The red line represents the mean of $y$ for any given $x$, called the _regression line_ or curve.
We see that with the chosen parameter values, x is quite informative about $y$, but not perfectly so:
The data are scattered around the line, with the amount of scatter being determined by $\sigma$. 

### Exercise

1. Play around with different parameter values for $\alpha$, $\beta$, and $\sigma$ to get a feeling for how the model works.

```{r}

```

2. Set $\beta$ and $\sigma$ to zero and think about what they imply, respectively.

```{r}

```

## Predictions

## Inference

## Model fit

## Numerical and categorical predictors


<!-- # Multiple linear regression -->
<!---->
<!-- ## One binary and one numerical predictor -->
<!---->
<!-- ## Two numerical predictors -->
<!---->
<!-- ## More than two predictors -->
<!---->
<!-- ## Interaction effects -->



