---
title: "Quantitative Methods and Statistics"
subtitle: "An applied course using the `R` programming language"
author: 
  - name: "Jakob Hoffmann"
    affiliation: "Economic Geography Group, LMU Munich"
format:
  html:
    toc: true
---


# Experiments

```{r}
# Boston housing data

df <- MASS::Boston |>
  as_tibble() |>
  select(
    value = medv,
    rooms = rm,
    age = age,
    status = lstat,
    crime = crim
  )

# Munich rent data
df <- gamlss.data::rent |>
  as_tibble() |>
  select(rent = R, size = Fl, year = A, location = loc) |>
  mutate(location = factor(location))

# Class size project STAR
library(AER)
data(STAR)
df <- STAR |>
  as_tibble() |>
  select()
```

# Exam

## Problem 1: Load and wrangle data

The file `trade.csv` contains data about trade volume relative to GDP, obtained from Our World in Data (OWID).

1) Load the file into a data frame saved in a variable called `trade`.

```{r}

```

2) Inspect the data set. How many rows and columns does the data frame have?

```{r}

```

3) The data set contains non-country entities (such as whole-world aggregates).
Remove them using the fact that country codes all have exactly three digits.
In the resulting data frame every observation (row) in the data frame is a country-year.

```{r}

```

4) Keep only the code, year, and trade columns.

```{r}

```

5) Data before the year 2000 is only sparsely available. Keep only observations for the years 2000 to 2024.

```{r}

```

6) Write a pipeline which does the previous steps in one go.

```{r}

df1 <- trade |>
  select(code = Code, year = Year, trade = `Trade (% of GDP)`) |>
  filter(nchar(code) == 3)

```

7) There is a second dataset called `democracy.csv`, which contains OWIDs Electoral Democracy Index (EDI).
Apply the same steps that you applied to the trade data.

```{r}

df2 <- democ |>
  select(
    code = Code,
    year = Year,
    edi = `Electoral Democracy Index`,
    region = `World region according to OWID`
  ) |>
  filter(nchar(code) == 3)

df3 <- gdppc |>
  select(
    code = Code,
    year = Year,
    gdp = `GDP per capita`,
  ) |>
  filter(nchar(code) == 3)

```

8) Join the two datasets (using in inner join) on country and year so that you have the trade and democracy data in a single data frame.

```{r}

df <- df1 |> left_join(df3) |> left_join(df2)

```

9) Drop observations with missing country code or year.

10) Mutate factor variables


# Problem 2: Describe and visualize the data

The csv file called `prepared.csv` contains a dataset with all the steps from problem 1 already applied, as well as additional data on GDP.
Load it and continue with this version for the remainder of the tasks.
(HINT: You can clear your environment now to get rid of previous variables.)

1) Write a pipeline to compute the number of countries as well as the mean, the median, the standard deviation, and the variation coefficient
of the democracy index by world region for the year 2020. Sort the result in decending order by the mean EDI.

```{r}

df |>
  filter(year == 2020) |>
  group_by(region) |>
  summarize(n = n(), mean = mean(edi), median = median(edi), sd = sd(edi), vc = sd / mean) |>
  arrange(desc(mean))

```

2a) Give a short interpretation of the findings for each of the three indicators.

2b) Explain the divergence of the mean and the median in Europe.


3) Create a plot comparing the development of the EDI from 2000 to 2024 in Germany (DEU), the USA, and Hungary (HUN). 

```{r}

```

4) Create a plot showing the histogram of the EDI in each world region in a separate panel.

```{r}

```

5) Scatterplot, correlation

```{r}

```


# Problem 3: Model the data

```{r}

dat <- inner_join(
  df |>
    filter(year == 1980) |>
    select(code, region, edi1980 = edi, gdp1980 = gdp, trade1980 = trade),
  df |>
    filter(year == 2019) |>
    select(code, region, edi2019 = edi, gdp2019 = gdp, trade2019 = trade)
) |>
  mutate(
    edi_change = (edi2019 - edi1980) * 100,
    trade_change = (trade2019 - trade1980),
    gdp_change = (gdp2019 - gdp1980) / gdp2019 * 100
  )

dat <- inner_join(
    df |> filter(year == 2000) |> select(code, region, edi2000 = edi, gdp2000 = gdp, trade2000 = trade),
    df |> filter(year == 2019) |> select(code, region, edi2019 = edi, gdp2019 = gdp, trade2019 = trade)
) |> mutate(
  edi_change = (edi2019 - edi2000) * 100,
  trade_change = (trade2019 - trade2000),
  gdp_change = (gdp2019 - gdp2000) / gdp2019 * 100
)

ggplot(dat, aes(edi_change, trade_change )) +
  geom_point() +
  geom_smooth(method = "lm")

```


```{r}
# OWID trade openness

trade <- read_csv("https://ourworldindata.org/grapher/trade-as-share-of-gdp.csv")
democ <- read_csv("https://ourworldindata.org/grapher/electoral-democracy-index.csv")
gdppc <- read_csv("https://ourworldindata.org/grapher/gdp-per-capita-worldbank.csv")

df <- trade |>
  inner_join(democ, by = c("Entity", "Year", "Code")) |>
  inner_join(gdppc, by = c("Entity", "Year", "Code")) |>
  select(
    country = Entity,
    code = Code,
    year = Year,
    trade = `Trade (% of GDP)`,
    democracy = `Electoral Democracy Index`,
    gdp = `GDP per capita`
  )

dat <- df |>
  filter(year == 2023) |>
  left_join(
    df |> filter(year == 2000) |> select(country, gdp_2000 = gdp, democracy_2000 = democracy),
    by = "country"
  ) |>
  mutate(
    gdp_change = (gdp - gdp_2000) / gdp_2000 * 100,
    democracy_change = (democracy - democracy_2000) / democracy_2000 * 100
  )

```

```{r}
```

```{r}
```

## Problem 2: Describe data

## Problem 3: Model data


```{r}
fit <- lm(log(value) ~ log(age) + log(status), data = df)
summary(fit)
```

```{r}
```
