---
title: "Quantitative Methods and Statistics"
subtitle: "Exam - Winter Term 2025/26"
author:
  - name: "Jakob Hoffmann"
    affiliation: "Economic Geography Group, LMU Munich"
format:
  html:
    toc: true
---

# Introduction

A key measure of a country's integration into the global economy is **trade openness**, calculated as the sum of exports and imports divided by GDP (expressed as a percentage).
A central question in understanding globalization is whether more open economies tend to be richer. The examination of this question is the primary subject of this exam.

The exam has three parts with a sum of **30 points**:

1. Loading and wrangling data (10 points)

2. Describing and visualizing data (10 points)

3. Fitting a model to data (10 points)


# Problem 1: Load and wrangle data (10 points)

The file `trade_raw.csv` contains data about trade openness (trade as % of GDP) obtained from Our World in Data.
HINT: Don't forget to load the packages you need before starting with the analysis.

1) Load the file into a data frame called `trade_raw`. (0.5 points)

```{r}
library(tidyverse)
trade_raw <- read_csv("data/trade_raw.csv")
```


2) Inspect the data set. How many rows and columns does it have? What variables does it contain? (0.5 points)

```{r}
dim(trade_raw)
```


3) Write a piece of code that keeps only the columns for country name, country code, year, and trade openness and renames them to `country`, `code`, `year`, and `trade_openness`. (1 points)

```{r}
select(trade_raw, country = Entity, code = Code, year = Year, trade_openness = `Trade (% of GDP)` )
```


4) The dataset contains non-country entities (such as world aggregates and regions).
Write a piece of code that keeps only actual countries by using the fact that ISO country codes have exactly three characters. (1 points)

```{r}
filter(trade_raw, str_length(Code) == 3)
```


5) Write a piece of code that keeps only observations for the year 2019. (1 points)

```{r}
trade <- filter(trade_raw, Year == 2019)
```


6) Write a pipeline that performs the three previous steps in sequence. (2 points)

```{r}
trade <- trade_raw |>
  select(
    country = Entity,
    code = Code,
    year = Year,
    trade_openness = `Trade (% of GDP)`
  ) |>
  filter(str_length(code) == 3, year == 2019)
```


7) There is a second dataset called `gdp_raw.csv` containing GDP per capita data.
Load it and apply the same steps from 3 to 5 in one pipeline (make sure to name variables appropriately).
Save the result as a variable called `gdp`. (2 points)

```{r}
gdp <- read_csv("data/gdp_raw.csv") |>
  select(
    country = Entity,
    code = Code,
    year = Year,
    gdp = `GDP per capita`,
    region = `World region according to OWID`
  ) |>
  filter(str_length(code) == 3, year == 2019)
```


8) Combine the two datasets using an inner join. (2 Points)

```{r}
inner_join(trade, gdp)
```


# Problem 2: Describe and visualize the data (10 points)

The file `trade_gdp.csv` contains a pre-cleaned dataset with trade openness, GDP per capita, a variable with a democracy index, and world region for 153 countries in 2019.
Load this file and use it for the remaining problems.

```{r}
df <- read_csv("data/trade_gdp.csv")
```


1) GDP per capita is highly skewed.
Create a new variable `log_gdp_pc` containing the natural logarithm of GDP per capita.
Show the distribution of the resulting variable as a histogram. (2 points)

```{r}
df <- mutate(df, log_gdp_pc = log(gdp_per_capita))
```


2) Create a scatter plot showing the relationship between trade openness (x-axis) and log GDP per capita (y-axis).
Color the points by world region and describe the result in one sentence. (2 points)

```{r}
ggplot(df, aes(trade_openness, log_gdp_pc, color = region)) + geom_point()
```


3) Compute the Pearsson correlation coefficient between trade openness and log GDP per capita. Interpret the result in one sentence. (2 points)

```{r}
cor(df$trade_openness, df$log_gdp_pc)
```


4) Compute the mean, median, and standard deviation for trade openness by world region. Give a brief summary of the result. (2 points)

```{r}
group_by(df, region) |>
  summarize(
    mean = mean(trade_openness),
    median = median(trade_openness),
    sd = sd(trade_openness)
  )
```


5) Visualize the distribution of trade openness by world region (pick an adequate form of visualization). (2 points)

```{r}
ggplot(df, aes(region, trade_openness)) + geom_boxplot()
```


# Problem 3: Fit a model to the data (10 points)


1) Fit a simple linear regression model (using the `lm()` function) predicting log GDP per capita from trade openness. Save the model as `m1`. (1 points)

```{r}
m1 <- lm(log_gdp_pc ~ trade_openness, data = df)
```


2) Visualize the linear model estimated above against the scatterplot of the data.
Use the visualization to get a (rough) prediction of log GDP per capita for a country with a trade openness of 200%. (1 points)

```{r}
ggplot(df, aes(trade_openness, log_gdp_pc)) + geom_point() + geom_smooth(method = "lm")
```


3) Give an interpretation of the coefficient for trade openness. (1 points)

HINT: Remember that changes in log-scaled variables can be approximately interpreted as relative changes.

```{r}

```

4) Summarize the estimation uncertainty. Is the coefficient for trade openness greater than zero with high certainty?  (2 points)

```{r}
summary(m1)
```


5) The dataset also contains a **democracy index** (ranging from 0 to 1, where higher values indicate more democratic institutions).
Give a short discussion about why this might (not) be a confounder for the causal effect of trade openness on GDP per capita. (2 points)

```{r}

```


6) Estimate a multiple regression model controlling for the level of democratic institutions.
Give a brief interpretation of the coefficient of trade openness in this model. (2 points)

```{r}
m2 <- lm(log_gdp_pc ~ trade_openness + democracy, data = df)
```


7) Compare this model to the first one and summarize the differences.
How did adjusting for the democracy index change the estimated effect of trade openness? (1 points)

```{r}
m1
m2
```

